Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`

### 2021-03-18 manuscript.Rmd
not even sure what has changed

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof].

In this work, I used `bcftools` and `freebayes`callers to find biallelic SNPs. The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies. Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015], though others have reported high concordance and accuracy in their results [@hwang2015]. Best method could depend on nature of the sequencing data: eg. whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.

>>>>

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof]. Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015]. However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data: (whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).
